---
title: "Exogenous Variables"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exogenous Variables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r setup, include=FALSE}
library(httptest2)
.mockPaths("../tests/mocks")
start_vignette(dir = "../tests/mocks")

original_options <- options("NIXTLA_API_KEY"="dummy_api_key", digits=7)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  fig.width = 7, 
  fig.height = 4
)
```

```{r}
library(nixtlar)
```

## 1. Exogenous variables 

Exogenous variables are external factors that provide additional information about the behavior of the target variable in time series forecasting. These variables, which are correlated with the target, can significantly improve predictions. Examples of exogenous variables include weather data, economic indicators, holiday markers, and promotional sales.

`TimeGPT` allows you to include exogenous variables when generating a forecast. This vignette will show you how to include them. It assumes you have already set up your API key. If you haven't done this, please read the [Get Started](https://nixtla.github.io/nixtlar/articles/anomaly-detection.html) vignette first. 

## 2. Load data 

For this vignette, we will use the electricity consumption dataset with exogenous variables included in `nixtlar`. This dataset contains hourly prices from five different electricity markets, along with two exogenous variables related to the prices and binary variables indicating the day of the week.

````{r}
df_exo_vars <- nixtlar::electricity_exo_vars
head(df_exo_vars)
````

When using exogenous variables, you must provide their future values to cover the complete forecast horizon; otherwise, `TimeGPT` will result in an error. Ensure that the dates of the future exogenous variables exactly match the forecast horizon. For the electricity consumption dataset with exogenous variables, `nixtlar` provides their values for the next 24 steps ahead.

````{r}
future_exo_vars <- nixtlar::electricity_future_exo_vars
head(future_exo_vars)
````

## 3. Forecast with exogenous variables 

To generate a forecast with exogenous variables, use the `nixtla_client_forecast` function as you would for forecasts without them. The only difference is that you must add the exogenous variables using the `X_df` argument.

Keep in mind that the default names for the time and target columns are `ds` and `y`, respectively. If your time and target columns have different names, specify them with `time_col` and `target_col`. Since this dataset has multiple ids (one for every electricity market), you will need to specify the name of the column that contains these ids, which in this case is `unique_id`. To do this, simply use `id_col="unique_id"`.  

````{r}
fcst_exo_vars <- nixtla_client_forecast(df_exo_vars, h = 24, id_col = "unique_id", X_df = future_exo_vars)
head(fcst_exo_vars)
````

For comparison, we will also generate a forecast without the exogenous variables. 

````{r}
df <- nixtlar::electricity # same dataset but without the exogenous variables

fcst <- nixtla_client_forecast(df, h = 24, id_col = "unique_id")
head(fcst)
````

## 4. Plot TimeGPT forecast

`nixtlar` includes a function to plot the historical data and any output from `nixtla_client_forecast`, `nixtla_client_historic`, `nixtla_client_anomaly_detection` and `nixtla_client_cross_validation`. If you have long series, you can use `max_insample_length` to only plot the last N historical values (the forecast will always be plotted in full). 

```{r}
nixtla_client_plot(df_exo_vars, fcst_exo_vars, id_col = "unique_id", max_insample_length = 500)
```

```{r, include=FALSE}
options(original_options)
end_vignette()
```
