---
title: "Historical Forecast"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Historical Forecast}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r setup, include=FALSE}
library(httptest2)
start_vignette("mocks")

options("TIMEGPT_TOKEN"="dummy_token")
options(digits=7)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  fig.width = 7, 
  fig.height = 4
)
```

```{r}
library(nixtlar)
```

## 1. TimeGPT Historical Forecast

```{r}
df <- nixtlar::electricity
head(df)
```

```{r}
timegpt_fitted_values <- timegpt_historic(df, id_col = "unique_id", level = c(80,95))
head(timegpt_fitted_values)
```

## 2. Plot TimeGPT forecast 
`nixtlar` includes a function to plot the historical data and any output from `timegpt_forecast`, `timegpt_historic`, `timegpt_anomaly_detection` and `timegpt_cross_validation`. If you have long series, you can use `max_insample_length` to only plot the last N historical values (the forecast will always be plotted in full). 

```{r}
timegpt_plot(df, timegpt_fitted_values, id_col = "unique_id", max_insample_length = 200)
```

```{r, include=FALSE}
end_vignette()
```




