---
title: "Prediction Intervals"
output: 
  rmarkdown::html_vignette:
    toc: true 
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Prediction Intervals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

---
```{r setup, include=FALSE}
library(httptest2)
.mockPaths("../tests/mocks")
start_vignette(dir = "../tests/mocks")

original_options <- options("NIXTLA_API_KEY"="dummy_api_key", digits=7)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  fig.width = 12, 
  fig.height = 8
)
```

```{r}
library(nixtlar)
```

## 1. Uncertainty quantification via prediction intervals  
For uncertainty quantification, `TimeGPT` can generate both prediction intervals and quantiles, offering a measure of the range of potential outcomes rather than just a single point forecast. In real-life scenarios, forecasting often requires considering multiple alternatives, not just one prediction. This vignette will explain how to use prediction intervals with `TimeGPT` via the `nixtlar` package.

A prediction interval is a range of values that the forecast can take with a given probability, often referred to as the confidence level. Hence, a 95% prediction interval should contain a range of values that includes the actual future value with a probability of 95%. Prediction intervals are part of probabilistic forecasting, which, unlike point forecasting, aims to generate the full forecast distribution instead of just the mean or the median of that distribution.

This vignette assumes you have already set up your API key. If you haven't done this, please read the [Get Started](https://nixtla.github.io/nixtlar/articles/get-started.html) vignette first. 

## 2. Load data 
For this vignette, we will use the electricity consumption dataset that is included in `nixtlar`, which contains the hourly prices of five different electricity markets. 

```{r}
df <- nixtlar::electricity
head(df)
```

## 3. Forecast with prediction intervals
`TimeGPT` can generate prediction intervals when using the following functions: 

```{r, eval=FALSE}
- nixtlar::nixtla_client_forecast()
- nixtlar::nixtla_client_historic() 
- nixtlar::nixtla_client_detect_anomalies()
- nixtlar::nixtla_client_cross_validation()
```

For any of these functions, simply set the `level` argument to the desired confidence level for the prediction intervals. Keep in mind that `level` should be a vector with numbers between 0 and 100. You can use either `quantiles` or `level` for uncertainty quantification, but not both. 

```{r}
fcst <- nixtla_client_forecast(df, h = 8, level=c(80,95))
head(fcst)
```

Note that the `level` argument in the `nixtlar::nixtla_client_detect_anomalies()` function only uses the maximum value when multiple values are provided. Therefore, setting `level = c(90, 95, 99)`, for example, is equivalent to setting `level = c(99)`, which is the default value.

```{r}
anomalies <- nixtla_client_detect_anomalies(df) # level=c(90,95,99)
head(anomalies) # only the 99% confidence level is used 
```

## 4. Plot prediction intervals 
`nixtlar` includes a function to plot the historical data and any output from `nixtlar::nixtla_client_forecast`, `nixtlar::nixtla_client_historic`, `nixtlar::nixtla_client_detect_anomalies` and `nixtlar::nixtla_client_cross_validation`. If you have long series, you can use `max_insample_length` to only plot the last N historical values (the forecast will always be plotted in full). 

When available, `nixtlar::nixtla_client_plot` will automatically plot the prediction intervals. 

```{r}
nixtla_client_plot(df, fcst, max_insample_length = 100)
```

```{r}
nixtlar::nixtla_client_plot(df, anomalies, plot_anomalies = TRUE)
```

```{r, include=FALSE}
options(original_options)
end_vignette()
```


