---
title: "Special Topics"
output: 
  rmarkdown::html_vignette:
    toc: true 
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Special Topics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r setup, include=FALSE}
library(httptest2)
.mockPaths("../tests/mocks")
start_vignette(dir = "../tests/mocks")

original_options <- options("NIXTLA_API_KEY"="dummy_api_key", digits=7)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  fig.width = 7, 
  fig.height = 4
)
```

```{r}
library(nixtlar)
```

## Special topics 

This vignette explains some special topics regarding the use of `TimeGPT` via `nixtlar`. 

## 1. Handling missing values  

Before using `TimeGPT`, you need to ensure that: 

1. The target column contains no missing values (`NA`).  
2. Given the frequency of the data, the dates are continuous, with no missing dates between the start and the end dates.  

Regarding the second point, it is worth mentioning that it is possible to have multiple time series that start and end on different dates, but each series must contain uninterrupted data for its given time frame.

There are several ways to check for missing values in R. One method is with the `any` and `is.na` functions from base R. 

```{r}

df <- nixtlar::electricity # load data 

# create some missing values at random 
index <- sample(nrow(df), 10)
df$y[index] <- NA

# check for missing values 
any(is.na(df)) # will return TRUE if there are missing values 
```

If you find missing values in your data, you need to decide how to fill them, which is very context-dependent. For example, if you are dealing with daily retail data, a missing value most likely indicates that there were no sales on that day, and you can probably fill it with zero. However, if you are working with hourly temperature data, a missing value likely means that the sensor was not functioning correctly, and you might prefer to use interpolation to fill the missing values. Whatever you decide to do, always keep in mind the nature of your data.

Checking if there are missing dates is more complicated since it depends on the frequency of the data. Sometimes plotting can help spot large gaps. `nixtlar` has a plotting function called `nixtla_client_plot` that can be used for this. 

However, this method is ineffective when the missing dates are not continuous. One possible solution is to compare the dates for every unique id with a vector of dates generated using the start date, the end date, and the frequency of your data. This requires knowing such information, which can become tricky when working with hundreds or thousands of time series.

## 2. Specifying the frequency of your data 

The frequency parameter is crucial when working with time series data because it informs the model about the expected intervals between data points. The core functions of `nixtlar` that interface with `TimeGPT`, such as `nixtla_client_forecast`, `nixtla_client_historic`, `nixtla_client_detect_anomalies`, and `nixtla_client_cross_validation`, require you to specify the `freq` parameter, although in some cases `nixtlar` can deduce it from your data. 

`TimeGPT` supports the following aliases: 

| Frequency                 | Alias        | 
|---------------------------|--------------|
| Yearly                    | Y            | 
| Quarterly                 | Q, QS, or QE |
| Monthly                   | M, MS, or MS | 
| Weekly (starting Sundays) | W            | 
| Daily                     | d            |
| Hourly                    | h            | 
| Minute-level              | min          |
| Second-level              | s            |
| Business day              | B            |

In this table, QS and MS stand for quarter and month start, while QE and ME stand for quarter and month end. Hourly and subhourly frequencies can be preceded by an integer, such as "6h", "10min" or "30s". **Only the aliases "min" and "S" are allowed for minute and second-level frequencies**. 

The default value of the frequency parameter is `NULL`. When this parameter is not specified, `nixtlar` will attempt to determine the frequency of your data. 

```{r}

df <- nixtlar::electricity
fcst <- nixtlar::nixtla_client_forecast(df, h = 8, level = c(80,95)) # freq = "h"
# infer the frequency when `freq` is not specified 
```

**Currently, `nixtlar` can't infer business day frequency, so you must set it directly using `freq="B"`.** 

```{r, include=FALSE}
options(original_options)
end_vignette()
```

